\framework: Morphology-based Entity and Relational Entity Extraction Framework for Arabic

### 2018-06-23 tex/merf/TALMajor/results.tex
format issues ficed: figure and table caption style
minor changes

\major{
In this section we evaluate \framework with four case studies. 
We perform a survey like evaluation where developers manually 
built task specific information extraction tools for the case studies 
and other developers built equivalent \framework tools. 
The aim of the comparison is to showcase that \framework enables 
fast development of linguistic applications with similar accuracy 
and a reasonable affordable overhead in computational time. 
We report development time,
size of developed code versus size of grammar,
running time, and 
precision-recall as metrics
of cost, complexity, overhead, and accuracy,
respectively.

@@

We evaluated ANGE with
Musnad Ahmad, a hadith book, where we constructed an annotated 
golden reference containing 1,865 words.
We evaluated ATEEMA 
with articles from issues of  the Lebanese 
Al-Akhbar newspaper where we constructed 
an annotated golden reference containing 1,677 words. 
For the genealogical tree extraction we used 
an extract from the Genesis biblical text with 1,227 words.
Finally, we used an annotated article from the Lebanese Assafir
newspaper with 1,399 words to evaluate the NUMNORM case study
\footnote{available at \url{http://www.assafir.com} and 
\url{http://www.al-akhbar.com}.}. 
%
In the online appendix%
~\footnote{available at ~\url{-omitted-for-anonymity-}}%http://webfea.fea.aub.edu.lb/fadi/pdfs/merfappendix.pdf
, we report on eight additional \framework case studies.
Manual annotators inspected the outcome and provided corrections where tools 
made mistakes.
The corrections form the manual gold annotation that we compared against.
}

% Table generated by Excel2LaTeX from sheet 'Sheet1'
\begin{table}[tb!]
  \centering
  \caption{\framework compared to task specific applications.}
  \label{tab:results}%
  \resizebox{\columnwidth}{!}{
    \begin{tabular}{l|l|l|l|ll|l}
     \toprule
      \multirow{2}{*}{Task} & Size & Development& Run & \multicolumn{2}{c|}{Accuracy} & \multirow{2}{*}{Ease of Composition}\\
      & (words) & time & time(s) & Recall & \multicolumn{1}{c|}{Precision} & \\
    \midrule
    %ANGE~\cite{ZaMaFlairs2012HadithBio}
      ANGE & 1,865  
    & 2 months & 1.79 & 0.99 & 0.99 & 3000+ lines of code\\
      \framework &  & 3 hours & 7.24 & 0.99 & 0.93  & 8 MBFs and 4 MREs\\
    \midrule
    %ATEEMA~\cite{ZaMa2012IJCLATime}
      ATEEMA & 1,677 & 1.5 months & 2.53 & 0.88  & 0.89  & 1000+ lines of code  \\
      \framework & & 3 hours & 3.14 & 0.91  & 0.81  & 3 MBFs and 2 MREs\\
    \midrule
    %Genealogy tree~\cite{ZaMaHaCicling2012Entity} 
      Genealogy tree & 1,227 
    & 3 weeks & 0.74 & 0.96 & 0.98 & 3000+ lines of code\\
      \framework &  & 4 hours & 2.28 & 0.84 & 0.93  & 3 MBFs and 3 MREs\\
    \midrule
      NUMNORM &  1,399 & 1 week & 0.32 & 0.91 & 0.93  & 500 lines of code\\
      \framework &  & 1 hour & 1.53 & 0.91 & 0.90 & 3 MBFs/1 MRE/57 lines\\
    \bottomrule
    \end{tabular}%
    }
\vspace{-1em}
\end{table}%

Table~\ref{tab:results} reports the development time,
extraction runtime, recall and precision 
of the output MRE tags, 
the size of the task in lines of code or in number of \framework rules, 
for both the standalone task specific and the \framework implementations.
\major{
The development time measures the time required for 
developing the case study.
For instance, ANGE~\cite{ZaMaFlairs2012HadithBio} required two 
months of development by a research assistant with six and 14 hours of 
course work and teaching duties, respectively.
Recall refers to the fraction of the entities correctly detected against the
total number of entities. 
Precision refers to the fraction of correctly
detected entities against the total number of extracted entities. 
}

\major{
Table~\ref{tab:results} provides runtime results of \framework 
compared to the task specific implementations while running 
MBF and MRE simulations jointly.
This is a rough estimate of the complexity of the \framework simulator. 
%
The complexity of the MBF simulation is the total number of morphological 
solutions for all the words multiplied by the number of user-defined MBFs.
We do not provide a limit on the number of user defined formulae.
In practice, we did not encounter more than ten formulae per case study.
As for the complexity of MRE simulation, converting the rules into 
non-deterministic finite state machines (NDFSM) is done once. 
Simulating an NDFSM over the MBF tags is potentially exponential. 
In practice all our case studies terminated within a predetermined 
time bound of less than 30 minutes. 
\framework required reasonably more runtime than the task specific 
implementations and reported acceptable and 
slightly less precision metrics with around
the same recall.
}

@@

\setcode{utf8}
\setarab
\begin{table}[tb!]
  \centering
  \caption{Narrator chain example.}
  \begin{Verbatim}[xleftmargin=1.5cm,fontsize=\relsize{-1},commandchars=\\\{\},codes={\catcode`$=3 \catcode`_=8}]
name:   PN ((MEAN)? PN)*;
nar:    name ((NONE)^3 FAM (NONE)^3 name)*;
pbuh:   BLESS GOD UPONHIM GREET;
nchain: ($s_1=$TOLD $s_2=$nar)+ ((PN|FAM|NONE)^8 pbuh)?
\end{Verbatim}
\resizebox{0.8\columnwidth}{!}{
    \begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}
    \toprule 
    \notrrl{القعقاع} & \notrrl{بن} & \notrrl{عمارة} & \notrrl{عن} & \notrrl{جرير} & \notrrl{حدثنا} & \notrrl{سعيد} & \notrrl{بن} & \notrrl{قتيبة} & \notrrl{حدثنا} \\
    \midrule 
    \noarrl{القعقاع} & \noarrl{بن} & \noarrl{عمارة} & \noarrl{عن} & \noarrl{جرير} & \noarrl{حدثنا} & \noarrl{سعيد} & \noarrl{بن} & \noarrl{قتيبة} & \noarrl{حدثنا} \\
    \midrule
    PN    & FAM   & PN    & TOLD  & PN    & TOLD  & PN    & FAM   & PN    & TOLD \\
    \midrule
    name & & name & & name & & name &  & name & \\
    \midrule
    \multicolumn{3}{|c|}{nar} &       & nar   &       & \multicolumn{3}{c|}{nar} &  \\
    \midrule
    \multicolumn{10}{|c|}{nchain}
    \\
    \bottomrule
    \end{tabular}%
}
  \label{tab:nchain}%
  \vspace{-1.5em}
\end{table}%
\setcode{standard}

@@

\begin{table}[tb!]
  \centering
  \caption{\framework MBF and user-defined relation accuracy }
  \resizebox{0.65\columnwidth}{!}{
    \begin{tabular}{l|c|c|c|c}
     \toprule
     \multirow{2}{*}{Task} & \multicolumn{2}{c|}{MBF accuracy} & \multicolumn{2}{c}{relation accuracy}\\
     & Recall & Precision & Recall & Precision \\
    \midrule
    Narrator chain & 0.99 & 0.85 & 0.99 & 0.98 \\
    Number normalization & 0.99 & 0.99 & 0.97 & 0.95 \\
    Temporal entity & 0.99 & 0.52 & 0.98 & 0.89 \\
    Genealogy tree & 0.99 & 0.75 & 0.81 & 0.96 \\
    \bottomrule
    \end{tabular}%
    }
  \label{tab:mbfer}%
  \vspace{-1.5em}
\end{table}%

@@

\vspace{-2em}
\subsection{Genealogy tree}
\vspace{-1em}

@@

\newcommand*{\fvtextcolor}[2]{\textcolor{#1}{#2}}
\begin{figure}[tb!]
\centering
  \begin{tabular}{p{5.5cm}p{5.8cm}}
\begin{Verbatim}[fontsize=\relsize{-1},frame=single,label=TMB algorithm,commandchars=\\\[\]] 
cout << \fvtextcolor[red][$s1.text];
if(isHundred) {
  if(current != 0) {
    previous += current;
  }
  current = currentH * \fvtextcolor[red][$s1.number];
  currentH = 0;
  isHundred = false;
  isKey = true;
} else if(current == 0) {
  current = \fvtextcolor[red][$s1.number];
  isKey = true;
} else if(!isKey) {
  isKey = true;
  current = current * \fvtextcolor[red][$s1.number];
} else {
  previous += current;
  current = \fvtextcolor[red][$s1.number];}
\end{Verbatim}
&
\begin{Verbatim}[fontsize=\relsize{-1},frame=single,label=DT algorithm,commandchars=\\\[\]] 
if(isHundred) {currentH += \fvtextcolor[red][$s0.number];
} else if(current == 0) {
  current = \fvtextcolor[red][$s0.number];
} else if(isKey) {
  previous += current;
  current = \fvtextcolor[red][$s0.number];
} else {current += \fvtextcolor[red][$s0.number]; }
isKey = false;
\end{Verbatim}
\begin{Verbatim}[fontsize=\relsize{-1},frame=single,label=H algorithm,commandchars=\\\[\]] 
isHundred = true;
if(current == 0)  {
  currentH = \fvtextcolor[red][$s2.number];
} else if(!isKey) {
  currentH = current * \fvtextcolor[red][$s2.number];
  current = 0;
} else {currentH = \fvtextcolor[red][$s2.number];}
isKey = false;
\end{Verbatim}
\\ 
\end{tabular}
  \vspace{-2em}
\caption{Actions for TMB, DT, and H MRE expressions.}
  \vspace{-1.5em}
\label{fig:numnormalgo}
\end{figure}

@@

We implemented a number normalization extractor using \framework and 
compared it with {\em NUMNORM}, a 
C++ implementation for number normalization. 
First, we defined the MBFs \cci{DT}, \cci{H}, and \cci{TMB}
to denote (1) digits and tens, (2) hundreds, and (3) 
thousands, millions, and billions, respectively.
The \cci{num} MRE 
\cci{(DT|TMB|H)+} is one or more \cci{DT}, \cci{TMB}, or \cci{H} tags. 
\framework assigns the symbols $s_1$, $s_2$, and $s_3$ 
for the sub-expressions \cci{DT}, \cci{TMB}, and \cci{H}, respectively. 
%The actions associated with the sub-expressions \cci{DT}, \cci{TMB}, and \cci{H} are presented in the appendix.
\major{
Figure~\ref{fig:numnormalgo} shows the actions associated with the \cci{DT}, \cci{TMB}, and \cci{H} subexpressions that cumulatively compute the numeric value of the numeric expression match.
The actions use \framework API to access features of the matches such 
as text (\cci{\$s1.text}) and numeric 
value (\cci{\$s1.number}) of literal numbers such as numbers from one to ten.
}
%
Table~\ref{tab:mbfer} shows high accuracy in MBF tagging and relation extraction 
with 99\% and 97\% recall and 99\% and 95\% precision, respectively.

>>>>

In this section we evaluate \framework with four case studies. 
We perform a survey like evaluation where developers manually 
built task specific information extraction tools for the case studies 
and other developers built equivalent \framework tools. 
The aim of the comparison is to showcase that \framework enables 
fast development of linguistic applications with similar accuracy 
and a reasonable affordable overhead in computational time. 
We report development time,
size of developed code versus size of grammar,
running time, and 
precision-recall as metrics
of cost, complexity, overhead, and accuracy,
respectively.

@@

We evaluated ANGE with
Musnad Ahmad, a hadith book, where we constructed an annotated 
golden reference containing 1,865 words.
We evaluated ATEEMA 
with articles from issues of  the Lebanese 
Al-Akhbar newspaper where we constructed 
an annotated golden reference containing 1,677 words. 
For the genealogical tree extraction we used 
an extract from the Genesis biblical text with 1,227 words.
Finally, we used an annotated article from the Lebanese Assafir
newspaper with 1,399 words to evaluate the NUMNORM case study
\footnote{available at \url{http://www.assafir.com} and 
\url{http://www.al-akhbar.com}.}. 
%
In the online appendix%
~\footnote{available at ~\url{-omitted-for-anonymity-}}%http://webfea.fea.aub.edu.lb/fadi/pdfs/merfappendix.pdf
, we report on eight additional \framework case studies.
Manual annotators inspected the outcome and provided corrections where tools 
made mistakes.
The corrections form the manual gold annotation that we compared against.

\begin{table}[tb]
  \centering
  \resizebox{\columnwidth}{!}{
    \begin{tabular}{l|l|l|l|ll|l}
     \toprule
      \multirow{2}{*}{Task} & Size & Development& Run & \multicolumn{2}{c|}{Accuracy} & \multirow{2}{*}{Ease of Composition}\\
      & (words) & time & time(s) & Recall & \multicolumn{1}{c|}{Precision} & \\
    \midrule
    %ANGE~\cite{ZaMaFlairs2012HadithBio}
      ANGE & 1,865  
    & 2 months & 1.79 & 0.99 & 0.99 & 3000+ lines of code\\
      \framework &  & 3 hours & 7.24 & 0.99 & 0.93  & 8 MBFs and 4 MREs\\
    \midrule
    %ATEEMA~\cite{ZaMa2012IJCLATime}
      ATEEMA & 1,677 & 1.5 months & 2.53 & 0.88  & 0.89  & 1000+ lines of code  \\
      \framework & & 3 hours & 3.14 & 0.91  & 0.81  & 3 MBFs and 2 MREs\\
    \midrule
    %Genealogy tree~\cite{ZaMaHaCicling2012Entity} 
      Genealogy tree & 1,227 
    & 3 weeks & 0.74 & 0.96 & 0.98 & 3000+ lines of code\\
      \framework &  & 4 hours & 2.28 & 0.84 & 0.93  & 3 MBFs and 3 MREs\\
    \midrule
      NUMNORM &  1,399 & 1 week & 0.32 & 0.91 & 0.93  & 500 lines of code\\
      \framework &  & 1 hour & 1.53 & 0.91 & 0.90 & 3 MBFs/1 MRE/57 lines\\
    \bottomrule
    \end{tabular}%
    }
   \caption{\label{tab:results}\framework compared to task specific applications.}
%\vspace{-1em}
\end{table}%

Table~\ref{tab:results} reports the development time,
extraction runtime, recall and precision 
of the output MRE tags, 
the size of the task in lines of code or in number of \framework rules, 
for both the standalone task specific and the \framework implementations.
The development time measures the time required for 
developing the case study.
For instance, ANGE~\cite{ZaMaFlairs2012HadithBio} required two 
months of development by a research assistant with six and 14 hours of 
course work and teaching duties, respectively.
Recall refers to the fraction of the entities correctly detected against the
total number of entities. 
Precision refers to the fraction of correctly
detected entities against the total number of extracted entities.

Table~\ref{tab:results} provides runtime results of \framework 
compared to the task specific implementations while running 
MBF and MRE simulations jointly.
This is a rough estimate of the complexity of the \framework simulator. 
%
The complexity of the MBF simulation is the total number of morphological 
solutions for all the words multiplied by the number of user-defined MBFs.
We do not provide a limit on the number of user defined formulae.
In practice, we did not encounter more than ten formulae per case study.
As for the complexity of MRE simulation, converting the rules into 
non-deterministic finite state machines (NDFSM) is done once. 
Simulating an NDFSM over the MBF tags is potentially exponential. 
In practice all our case studies terminated within a predetermined 
time bound of less than 30 minutes. 
\framework required reasonably more runtime than the task specific 
implementations and reported acceptable and 
slightly less precision metrics with around
the same recall.

@@

\setcode{utf8}
\setarab
\begin{table}[tb!]
  \centering
  \begin{Verbatim}[xleftmargin=1.5cm,fontsize=\relsize{-1},commandchars=\\\{\},codes={\catcode`$=3 \catcode`_=8}]
name:   PN ((MEAN)? PN)*;
nar:    name ((NONE)^3 FAM (NONE)^3 name)*;
pbuh:   BLESS GOD UPONHIM GREET;
nchain: ($s_1=$TOLD $s_2=$nar)+ ((PN|FAM|NONE)^8 pbuh)?
\end{Verbatim}
\resizebox{0.8\columnwidth}{!}{
    \begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}
    \toprule 
    \notrrl{القعقاع} & \notrrl{بن} & \notrrl{عمارة} & \notrrl{عن} & \notrrl{جرير} & \notrrl{حدثنا} & \notrrl{سعيد} & \notrrl{بن} & \notrrl{قتيبة} & \notrrl{حدثنا} \\
    \midrule 
    \noarrl{القعقاع} & \noarrl{بن} & \noarrl{عمارة} & \noarrl{عن} & \noarrl{جرير} & \noarrl{حدثنا} & \noarrl{سعيد} & \noarrl{بن} & \noarrl{قتيبة} & \noarrl{حدثنا} \\
    \midrule
    PN    & FAM   & PN    & TOLD  & PN    & TOLD  & PN    & FAM   & PN    & TOLD \\
    \midrule
    name & & name & & name & & name &  & name & \\
    \midrule
    \multicolumn{3}{|c|}{nar} &       & nar   &       & \multicolumn{3}{c|}{nar} &  \\
    \midrule
    \multicolumn{10}{|c|}{nchain}
    \\
    \bottomrule
    \end{tabular}%
}
\caption{\label{tab:nchain}Narrator chain example.}
 \vspace{-1em}
\end{table}
\setcode{standard}

@@

\begin{table}[tb]
  \centering
  \resizebox{0.8\columnwidth}{!}{
    \begin{tabular}{l|c|c|c|c}
     \toprule
     \multirow{2}{*}{Task} & \multicolumn{2}{c|}{MBF accuracy} & \multicolumn{2}{c}{relation accuracy}\\
     & Recall & Precision & Recall & Precision \\
    \midrule
    Narrator chain & 0.99 & 0.85 & 0.99 & 0.98 \\
    Number normalization & 0.99 & 0.99 & 0.97 & 0.95 \\
    Temporal entity & 0.99 & 0.52 & 0.98 & 0.89 \\
    Genealogy tree & 0.99 & 0.75 & 0.81 & 0.96 \\
    \bottomrule
    \end{tabular}%
    }
  \caption{\label{tab:mbfer}\framework MBF and user-defined relation accuracy.}
 \vspace{-1.5em}
\end{table}%

@@

\vspace{-1.5em}
\subsection{Genealogy tree}
\vspace{-1em}

@@

@@

\newcommand*{\fvtextcolor}[2]{\textcolor{#1}{#2}}
\begin{figure}[tb]
\centering
  \begin{tabular}{p{5.5cm}p{5.8cm}}
\begin{Verbatim}[fontsize=\relsize{-1},frame=single,label=TMB algorithm,commandchars=\\\[\]] 
cout << \fvtextcolor[red][$s1.text];
if(isHundred) {
  if(current != 0) {
    previous += current;
  }
  current = currentH * \fvtextcolor[red][$s1.number];
  currentH = 0;
  isHundred = false;
  isKey = true;
} else if(current == 0) {
  current = \fvtextcolor[red][$s1.number];
  isKey = true;
} else if(!isKey) {
  isKey = true;
  current = current * \fvtextcolor[red][$s1.number];
} else {
  previous += current;
  current = \fvtextcolor[red][$s1.number];}
\end{Verbatim}
&
\begin{Verbatim}[fontsize=\relsize{-1},frame=single,label=DT algorithm,commandchars=\\\[\]] 
if(isHundred) {currentH += \fvtextcolor[red][$s0.number];
} else if(current == 0) {
  current = \fvtextcolor[red][$s0.number];
} else if(isKey) {
  previous += current;
  current = \fvtextcolor[red][$s0.number];
} else {current += \fvtextcolor[red][$s0.number]; }
isKey = false;
\end{Verbatim}
\begin{Verbatim}[fontsize=\relsize{-1},frame=single,label=H algorithm,commandchars=\\\[\]] 
isHundred = true;
if(current == 0)  {
  currentH = \fvtextcolor[red][$s2.number];
} else if(!isKey) {
  currentH = current * \fvtextcolor[red][$s2.number];
  current = 0;
} else {currentH = \fvtextcolor[red][$s2.number];}
isKey = false;
\end{Verbatim}
\\ 
\end{tabular}
 \vspace{-2em}
\caption{Actions for TMB, DT, and H MRE expressions.}
\vspace{-1.5em}
\label{fig:numnormalgo}
\end{figure}

We implemented a number normalization extractor using \framework and 
compared it with {\em NUMNORM}, a 
C++ implementation for number normalization. 
First, we defined the MBFs \cci{DT}, \cci{H}, and \cci{TMB}
to denote (1) digits and tens, (2) hundreds, and (3) 
thousands, millions, and billions, respectively.
The \cci{num} MRE 
\cci{(DT|TMB|H)+} is one or more \cci{DT}, \cci{TMB}, or \cci{H} tags. 
\framework assigns the symbols $s_1$, $s_2$, and $s_3$ 
for the sub-expressions \cci{DT}, \cci{TMB}, and \cci{H}, respectively. 
%The actions associated with the sub-expressions \cci{DT}, \cci{TMB}, and \cci{H} are presented in the appendix.
Figure~\ref{fig:numnormalgo} shows the actions associated with the \cci{DT}, \cci{TMB}, and \cci{H} subexpressions that cumulatively compute the numeric value of the numeric expression match.
The actions use \framework API to access features of the matches such 
as text (\cci{\$s1.text}) and numeric 
value (\cci{\$s1.number}) of literal numbers such as numbers from one to ten.
%
Table~\ref{tab:mbfer} shows high accuracy in MBF tagging and relation extraction 
with 99\% and 97\% recall and 99\% and 95\% precision, respectively.

### 2018-06-23 tex/merf/TALMajor/sarf.tex
format issues ficed: figure and table caption style
minor changes

%~\footnote{In this document, we use the default ArabTeX transliteration style ZDMG.}
%{\bf Morphological analyzer (Sarf).~}
\major{
Morphological analysis is key to Arabic 
NLP
due to the exceptional degree of ambiguity in writing, 
the rich morphology, and the complex word derivation 
system~\cite{arabicmorph,shahrour2016camelparser,pasha2014madamira}.
}
Short vowels, also known as diacritics, are typically omitted in Arabic text
and inferred by readers~\cite{habash2006arabic}. 
For example, the word \RL{bn}%
can be interpreted as \RL{bon} {\tt ``coffee''} with a {\em damma} diacritic on the 
letter \utfrl{بـ} or 
\vocalize \RL{bin} (son of) with a 
{\em kasra} diacritic on the letter \utfrl{بـ}.

@@

We denote by 
\ss,
\pp,
\xx,
\PP,
\GG, and 
\AC, the set of 
all stems,
prefixes,
suffixes,
POS,
gloss, 
and user defined category tags, respectively. 
Let $T=\langle t_1,t_2,\ldots,t_M\rangle$ be a set of Arabic words denoting the 
text documents. 
\framework uses Sarf to compute a set of morphological solutions $M(t)=\{m_1,m_2,\ldots,m_N\}$
for each word $t\in T$. 
%
\major{
Each morphological solution $m\in M(t)$ is a tuple of the form 
$\langle p,s,x,P,G,C\rangle \in \pp \times \ss \times \xx \times \PP \times \GG \times \AC$ 
where $p=p_1 \ldots p_{|p|}$, $x=x_1 \ldots x_{|x|}$, 
$P=P_{p_1}\ldots P_{p_{|p|}} P_s P_{x_1}\ldots P_{x_{|x|}}$,
$G=G_{p_1}\ldots G_{p_{|p|}} G_s G_{x_1}\ldots G_{x_{|x|}}$, and
$C=C_{p_1} \ldots C_{p_{|p|}} C_s C_{x_1}\ldots C_{x_{|x|}}$. 
$P_{p_i}, G_{p_i},$ and $C_{p_i},1\le i \le |p|$ are the POS, Gloss and 
category tags of prefix $p_i$. 
$P_{x_j}, G_{x_j},$ and $C_{x_j},1\le j \le |x|$ are the POS, Gloss and 
category tags of suffix $x_i$. 
$P_s, G_s,$ and $C_s$ are the POS, Gloss and 
category tags of stem $s$. 
%
Intuitively, 
$p,x,P,G$ and $C$ are concatenations of prefix, suffix, POS, Gloss and
Category values, respectively.
}

%\setarab
\begin{table}[tb!]
  \centering
  \caption{Sample solution vector for \utfrl{فَسَيَأْكُلها}.}
  \label{tab:samplesolution}%
  \resizebox{0.7\columnwidth}{!}{
    \begin{tabular}{|r|c|c|c|c|c|}
          & \multicolumn{3}{c|}{\textbf{Prefixes}} & \textbf{Stem} & \textbf{Suffix} \\
    \textbf{Data} & \utfrl{فَ} & \utfrl{سَ} & \utfrl{يَ} & \utfrl{أْكُل} & \utfrl{ها} \\
    \textbf{POS} & CONJ+ & FUT+ & IV3MS+ & VERB\_IMPERFECT & IVSUFF\_DO:3FS \\
    \textbf{Gloss} & and/so & will & he/it & eat/consume & it/them/her \\
    \textbf{index} & \multicolumn{3}{c|}{10} & 13 & 16 \\
    \textbf{length} & \multicolumn{3}{c|}{3} & 3 & 2 \\
    \end{tabular}%
    }
  \vspace{-2em}
\end{table}%

>>>>

%~\footnote{In this document, we use the default ArabTeX transliteration style ZDMG.}
%{\bf Morphological analyzer (Sarf).~}
Morphological analysis is key to Arabic 
NLP
due to the exceptional degree of ambiguity in writing, 
the rich morphology, and the complex word derivation 
system~\cite{arabicmorph,shahrour2016camelparser,pasha2014madamira}.
Short vowels, also known as diacritics, are typically omitted in Arabic text
and inferred by readers~\cite{habash2006arabic}. 
For example, the word \RL{bn}%
can be interpreted as \RL{bon} {\tt ``coffee''} with a {\em damma} diacritic on the 
letter \utfrl{بـ} or 
\vocalize \RL{bin} (son of) with a 
{\em kasra} diacritic on the letter \utfrl{بـ}.

@@

We denote by 
\ss,
\pp,
\xx,
\PP,
\GG, and 
\AC, the set of 
all stems,
prefixes,
suffixes,
POS,
gloss, 
and user defined category tags, respectively.
Let $T=\langle t_1,t_2,\ldots,t_M\rangle$ be a set of Arabic words denoting the 
text documents.
\framework uses Sarf to compute a set of morphological solutions $M(t)=\{m_1,m_2,\ldots,m_N\}$
for each word $t\in T$. 
%
Each morphological solution $m\in M(t)$ is a tuple of the form 
$\langle p,s,x,P,G,C\rangle \in \pp \times \ss \times \xx \times \PP \times \GG \times \AC$ 
where $p=p_1 \ldots p_{|p|}$, $x=x_1 \ldots x_{|x|}$, 
$P=P_{p_1}\ldots P_{p_{|p|}} P_s P_{x_1}\ldots P_{x_{|x|}}$,
$G=G_{p_1}\ldots G_{p_{|p|}} G_s G_{x_1}\ldots G_{x_{|x|}}$, and
$C=C_{p_1} \ldots C_{p_{|p|}} C_s C_{x_1}\ldots C_{x_{|x|}}$. 
$P_{p_i}, G_{p_i},$ and $C_{p_i},1\le i \le |p|$ are the POS, Gloss and 
category tags of prefix $p_i$. 
$P_{x_j}, G_{x_j},$ and $C_{x_j},1\le j \le |x|$ are the POS, Gloss and 
category tags of suffix $x_i$. 
$P_s, G_s,$ and $C_s$ are the POS, Gloss and 
category tags of stem $s$. 
%
Intuitively, 
$p,x,P,G$ and $C$ are concatenations of prefix, suffix, POS, Gloss and
Category values, respectively.

%\setarab
\begin{table}[tb]
  \centering
  \resizebox{0.85\columnwidth}{!}{
    \begin{tabular}{|r|c|c|c|c|c|}
          & \multicolumn{3}{c|}{\textbf{Prefixes}} & \textbf{Stem} & \textbf{Suffix} \\
    \textbf{Data} & \utfrl{فَ} & \utfrl{سَ} & \utfrl{يَ} & \utfrl{أْكُل} & \utfrl{ها} \\
    \textbf{POS} & CONJ+ & FUT+ & IV3MS+ & VERB\_IMPERFECT & IVSUFF\_DO:3FS \\
    \textbf{Gloss} & and/so & will & he/it & eat/consume & it/them/her \\
    \textbf{index} & \multicolumn{3}{c|}{10} & 13 & 16 \\
    \textbf{length} & \multicolumn{3}{c|}{3} & 3 & 2 \\
    \end{tabular}%
    }
    \vspace{-1em}
  \caption{\label{tab:samplesolution}Sample solution vector for \utfrl{فَسَيَأْكُلها}.}
\end{table}%
